---
title: "Prompt, Run, Verify"
subtitle: "Analysis walkthrough · Workshop 3: Using LLMs for Data Analysis in Public Health"
author: "Heather Mattie, PhD · Harvard T.H. Chan School of Public Health"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                      fig.width = 7, fig.height = 4.5, dpi = 150)
crimson <- "#A51C30"
```

```{css, echo=FALSE}
body { font-family: 'IBM Plex Sans', Arial, sans-serif; color: #2A1E21; background: #FCFAF8; font-size: 16px; line-height: 1.6; }
h1, h2, h3, .title { font-family: 'Source Serif 4', Georgia, serif; color: #2A1E21; }
h1.title { color: #A51C30; font-size: 40px; }
h1 { border-bottom: 3px solid #A51C30; padding-bottom: 6px; margin-top: 40px; font-size: 30px; }
.subtitle, .author, .date { color: #5F5256; }
a, a:visited { color: #A51C30; }
pre, code { font-family: 'JetBrains Mono', 'Courier New', monospace; }
pre { background: #F7ECEE; border: 1px solid #E8D3D7; border-radius: 8px; }
pre:not([class]) { background: #FFFFFF; }
blockquote { border-left: 4px solid #A51C30; background: #F7ECEE; color: #2A1E21; font-size: 16px; border-radius: 0 8px 8px 0; }
.list-group-item.active, .list-group-item.active:hover, .list-group-item.active:focus { background: #A51C30; border-color: #A51C30; }
.tocify .list-group-item { color: #2A1E21; }
.btn-default { border-color: #E8D3D7; color: #A51C30; }
.checkpoint { background: #FFFFFF; border: 1px solid #E8D3D7; border-left: 6px solid #A51C30; border-radius: 8px; padding: 12px 16px; margin: 16px 0; }
```

> **How to use this file.** Every code chunk below is the kind of code an LLM will
> happily write for you. The *text* between the chunks is the part the LLM can't do for
> you: noticing what is wrong. Run each chunk with the green arrow and read the
> output before moving on. Click **Knit** to turn the whole file into a report.
>
> **First time?** Put this file and `nhanes_workshop.csv` in the same folder, open the
> file in RStudio, and run this once in the Console:
> `install.packages(c("tidyverse", "gtsummary", "flextable", "broom", "rmarkdown"))`

```{r packages}
library(tidyverse)   # data import, wrangling, ggplot2
library(gtsummary)   # Table 1 and regression tables
library(flextable)   # export tables to Word / PowerPoint
library(broom)       # tidy model output
```

# 1. Import and first look

**Data:** 1,000 adults sampled from the public `NHANES` R package (2009–2012 survey
cycles), reshaped to look like a typical study extract. The codebook is in section 4 of the Student Guide.

```{r import}
dat_raw <- read_csv("nhanes_workshop.csv", show_col_types = FALSE)
glimpse(dat_raw)
```

**Checkpoint (check 2: right dimensions and types?).** 1,000 rows, 10 columns — good.
But `race` is stored as a number (`dbl`). It is a *category* coded 1–5. R does not
know that, and neither will the LLM unless you tell it.

```{r summary-raw}
summary(dat_raw)
```

**Checkpoint (check 3: plausible values?).** Mean BMI is about 37.6. That alone might
not set off alarms — but the **maximum is 999**. That's a missing-value code, not a
person. The mean was quietly inflated by 9 fake values.

```{r count-999}
dat_raw |> count(bmi == 999)
```

# 2. Clean — the step that needs *you*

These decisions come from the codebook and your domain knowledge, not from the LLM.

```{r clean}
dat <- dat_raw |>
  mutate(
    bmi  = na_if(bmi, 999),                       # 999 = missing (codebook)
    race = factor(race, levels = 1:5,
                  labels = c("White", "Black", "Mexican American",
                             "Other Hispanic", "Other/Multiracial")),
    sex  = factor(sex, levels = c("female", "male"),
                  labels = c("Female", "Male")),
    education = factor(education,
                       levels = c("8th Grade", "9 - 11th Grade", "High School",
                                  "Some College", "College Grad")),
    diabetes    = factor(diabetes,    levels = c("No", "Yes")),
    phys_active = factor(phys_active, levels = c("No", "Yes")),
    ever_smoker = factor(ever_smoker, levels = c("No", "Yes"))
  )

summary(dat$bmi)   # now plausible: max 62.8, 9 NA
```

# 3. Descriptive statistics

```{r descriptives}
dat |>
  group_by(sex) |>
  summarize(n        = n(),
            mean_age = mean(age),
            mean_bmi = mean(bmi, na.rm = TRUE),
            mean_sbp = mean(sbp, na.rm = TRUE),
            n_missing_sbp = sum(is.na(sbp)))
```

**Checkpoint.** Do the group sizes add to 1,000? Are missing values counted, not
silently dropped?

# 4. Visualizations

```{r fig-bmi-hist}
ggplot(dat, aes(x = bmi)) +
  geom_histogram(binwidth = 2, fill = crimson, color = "white") +
  labs(x = "BMI (kg/m2)", y = "Number of participants",
       title = "Distribution of BMI") +
  theme_minimal()
```

```{r fig-bmi-sbp}
ggplot(dat, aes(x = bmi, y = sbp)) +
  geom_point(alpha = 0.35, color = "#5F5256") +
  geom_smooth(method = "lm", se = TRUE, color = crimson, fill = "#E8D3D7") +
  facet_wrap(~ sex) +
  labs(x = "BMI (kg/m2)", y = "Systolic blood pressure (mmHg)",
       title = "BMI and systolic blood pressure, by sex") +
  theme_minimal()
```

Notice the warning about removed rows: R is telling you some rows had missing BMI or
SBP. **Read warnings; don't scroll past them.**

```{r fig-diabetes-activity}
dat |>
  count(phys_active, diabetes) |>
  group_by(phys_active) |>
  mutate(pct = n / sum(n)) |>
  filter(diabetes == "Yes") |>
  ggplot(aes(x = phys_active, y = pct)) +
  geom_col(fill = crimson, width = 0.6) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Physically active", y = "Diabetes prevalence",
       title = "Diabetes prevalence by physical activity") +
  theme_minimal()
```

# 5. Table 1

```{r table1}
table1 <- dat |>
  select(age, sex, race, education, bmi, sbp, phys_active, ever_smoker, diabetes) |>
  tbl_summary(
    by = diabetes,
    label = list(age ~ "Age (years)", sex ~ "Sex", race ~ "Race/ethnicity",
                 education ~ "Education", bmi ~ "BMI (kg/m2)",
                 sbp ~ "Systolic BP (mmHg)", phys_active ~ "Physically active",
                 ever_smoker ~ "Smoked 100+ cigarettes"),
    statistic = list(all_continuous() ~ "{mean} ({sd})"),
    missing_text = "Missing"
  ) |>
  add_overall() |>
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Diabetes**") |>
  bold_labels()

table1 |> as_flex_table()
```

**Checkpoint (domain check).** Should Table 1 include p-values? Many journals and
methodologists (e.g., CONSORT for trials) say no. The LLM will add them if you ask,
and often if you don't. That's a *judgment call*, not a coding question.

# 6. Regression models

## Linear regression: systolic BP

```{r lm-sbp}
fit_lm <- lm(sbp ~ bmi + age + sex + race, data = dat)

tbl_regression(fit_lm,
               label = list(bmi ~ "BMI (per 1 kg/m2)", age ~ "Age (per year)",
                            sex ~ "Sex", race ~ "Race/ethnicity")) |>
  add_glance_source_note(include = c(nobs, r.squared)) |>
  as_flex_table()
```

**Checkpoint (check 2 again).** Table 1 has 1,000 people. How many are in the model?

```{r nobs}
nobs(fit_lm)
```

Rows with missing BMI or SBP were dropped (complete-case analysis). That should be
reported in your methods, and it is a choice worth discussing with your advisor.

## Logistic regression: diabetes

```{r glm-diabetes}
fit_glm <- glm(diabetes ~ bmi + age + sex + phys_active,
               data = dat, family = binomial)

tbl_regression(fit_glm, exponentiate = TRUE,
               label = list(bmi ~ "BMI (per 1 kg/m2)", age ~ "Age (per year)",
                            sex ~ "Sex", phys_active ~ "Physically active")) |>
  as_flex_table()
```

**Checkpoint (interpretation).** These are odds ratios, not risk ratios. With about 9%
prevalence the two are close here, but the gap grows as an outcome gets more common, and
an LLM summary rarely checks. The data are cross-sectional: associations, not effects.
LLM summaries often blur both.

# 7. Check 4: test on a tiny example with a known answer

```{r mini-test}
test <- tibble(
  sex = c("F", "F", "F", "M", "M", "M"),
  sbp = c(120, 130, NA, 140, 150, 160)
)
# Expected by hand: F = 125, M = 150
test |> group_by(sex) |> summarize(mean_sbp = mean(sbp, na.rm = TRUE))
```

If this returned 83.3 for F, the code would be treating NA as 0. You found the bug on
6 rows instead of in your manuscript.

# 8. Text vs. numbers demo

These are the 25 values from the "Try it" exercise. Paste them into a chat **with code
execution turned off** and ask for the mean and standard deviation. Then compare with
what R computes below.

```{r numbers-demo}
set.seed(27)
bp25 <- dat |> filter(!is.na(sbp)) |> slice_sample(n = 25) |> pull(sbp)
cat(bp25, sep = ", ")
c(mean = mean(bp25), sd = sd(bp25))
```

The model predicts digits that *look like* a mean. Code computes one.

# 9. The no-upload path (restricted data)

When data can't leave approved systems, give the LLM the **structure**, not the
**data**. Example prompt:

> **Context:** I'm working in R with the tidyverse and gtsummary. My data frame `dat`
> has 1,000 rows. Columns: `age` (numeric, 20–80), `sex` (character: "female"/"male",
> lowercase), `race` (integer 1–5; 1=White, 2=Black, 3=Mexican American, 4=Other
> Hispanic, 5=Other/Multiracial), `bmi` (numeric; 999 = missing), `sbp` (numeric, some NA),
> `diabetes` ("No"/"Yes").
> **Objective:** Create a Table 1 comparing participants with and without diabetes,
> and fit a linear regression of `sbp` on `bmi`, `age`, `sex`, and `race`.
> **Format:** One commented R code chunk. First, write code that creates a fake
> 50-row data frame with this exact structure so I can test the code without real data.

Below is the kind of fake-data code the LLM returns. You develop against it, then run
the finished code on the real data **inside the approved environment**.

```{r synthetic}
set.seed(1)
fake <- tibble(
  age      = sample(20:80, 50, replace = TRUE),
  sex      = sample(c("female", "male"), 50, replace = TRUE),
  race     = sample(1:5, 50, replace = TRUE),
  bmi      = c(round(rnorm(48, 28, 6), 1), 999, 999),
  sbp      = c(round(rnorm(48, 120, 15)), NA, NA),
  diabetes = sample(c("No", "Yes"), 50, replace = TRUE, prob = c(.9, .1))
)
glimpse(fake)
```

Note what made this work: the prompt said "lowercase," "999 = missing," and "integer
1–5." Leave those out and the code will run and be wrong (see Spot the bug in section 5 of the Student Guide).

# 10. Exporting results

Same source, many outputs:

```{r export, eval=TRUE}
dir.create("output", showWarnings = FALSE)

# Table 1 -> Word and PowerPoint
table1 |> as_flex_table() |> save_as_docx(path = "output/table1.docx")
table1 |> as_flex_table() |> save_as_pptx(path = "output/table1.pptx")

# Regression table -> Word
tbl_regression(fit_lm) |> as_flex_table() |>
  save_as_docx(path = "output/regression_sbp.docx")

# Figure -> PNG (for slides) and PDF (for journals)
p <- ggplot(dat, aes(bmi, sbp)) + geom_point(alpha = .35, color = "#5F5256") +
  geom_smooth(method = "lm", color = crimson, fill = "#E8D3D7") + theme_minimal() +
  labs(x = "BMI (kg/m2)", y = "Systolic BP (mmHg)")
ggsave("output/bmi_sbp.png", p, width = 6, height = 4, dpi = 300)
ggsave("output/bmi_sbp.pdf", p, width = 6, height = 4)

# Clean data -> CSV
write_csv(dat, "output/nhanes_clean.csv")

list.files("output")
```

**Why this beats downloading a .docx from the chat window:** this file reruns from raw
data to final tables. When a reviewer asks you to drop 20 participants, you change one
line and knit again.

```{r session-info}
sessionInfo()
```
